A dynamic bioenergetic model for coral-\emph{Symbiodinium} symbioses and coral bleaching as an alternate stable state

### 2017-02-08 JTB_ms/JTB_ms.tex
more updates for revision

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a new module of photo-oxidative stress.} Of primary
  interest to coral biologists and ecologists is symbiosis dysfunction
  under environmental stress, resulting in coral ``bleaching''--the loss
  of algal symbionts from the association (Jokiel and Coles 1977).
  Bleaching is thought to be triggered by photo-oxidative stress in
  \emph{Symbiodinium} (Weis 2008), which has been modeled previously as
  a response to absolute external irradiance (Eynaud, Nisbet, and Muller
  2011). However, this response may also depend on self-shading by other
  symbionts, CO\textsubscript{2} availability at the site of
  photosynthesis, and other non-photochemical quenching mechanisms. We
  incorporate these features into a new photo-oxidative stress module
  linking overreduction of the photosynthetic light reactions to
  downstream impacts of photoinhibition and photodamage. Importantly,
  this formulation introduces a link between
  CO\textsubscript{2}-limitation of photosynthesis and bleaching, and
  potential synergistic roles for heterotrophy and nutrient availability
  in influencing bleaching responses (Wooldridge 2014b)
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB theory in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation precludes modeling the full life cycle of corals as
  reproduction, larval stages, and metamorphosis are not considered, but
  greatly reduces theoretical complexity and parameter numbers, which is
  advantageous given the relative paucity of data for corals. Moreover,
  dynamics of the symbiosis (i.e., changes in symbiont to host biomass,
  including bleaching and recovery) and coral biomass growth are
  efficiently captured with this simpler formulation, which also
  increases accessibility for biologists and ecologists without
  requiring significant expertise in DEB theory.
\item
  \emph{Provide well-documented, open-access code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide open access to the
  model in the form of detailed, commented code written in the R
  language (R Core Team 2014). With an accessible and modular framework,
  we envision this model as a resource for further development by the
  scientific community to include additional complexity and
  problem-specific components. We chose the R language, which is freely
  available and in common use by biologists and ecologists, to reach the
  widest possible audience with this work.
\end{enumerate}

With these as our primary motivations, we describe a simplified approach
to dynamic bioenergetic modeling of coral-\emph{Symbiodinium} symbioses
that dynamically integrates the influence of external irradiance,
nutrients, and prey availability on coral growth and symbiosis dynamics
(i.e., symbiont:host biomass ratios), allowing for the possibility of
coral bleaching in the event of photooxidative stress. In the following
sections, we describe and provide rationale for the model structure, and
demonstrate a range of steady state and dynamic behaviors that are
consistent with observed phenomena.

@@

In this dynamical system, both partners acquire and use carbon and
nitrogen to construct biomass: the symbiont fixes carbon through
photosynthesis, and receives nitrogen shared by the host, while the host
acquires nitrogen from the environment and receives carbon shared by the
symbiont. A graphical representation of the model is presented in Fig.
1, and each model flux and parameter is defined in Tables 1 and 2,
respectively. We use C-moles (C-mol) as the unit of biomass for
consistency with the rigorous mass balance of DEB theory: 1 C-mol is
equivalent to the amount of biomass containing 1 mole of carbon atoms.
Host biomass (expressed as C-mol H), symbiont biomass (expressed as
C-mol S), and prey biomass (expressed as C-mol X) have fixed, but
different, molar N:C ratios (Table 2). Biomass is produced from carbon
and nitrogen by synthesizing units (SU), which are mathematical
specifications of the formation of a product from two substrates; we use
the ``parallel complementary'' formulation of Kooijman (2010) (page 105,
Fig. 3.7) to specify these fluxes. The two state variables of this
system are symbiont biomass and coral biomass; because resources are
acquired proportionally to surface area, and surface area is
proportional to volume, biomass increases exponentially during growth
(indeed, corals grow exponentially (Bak 1976)).

Environmental stress is implemented in the form of photooxidative
stress, which is thought to be a primary trigger of coral bleaching
(Lesser 1997; Weis 2008; Wooldridge 2009). To simulate bleaching, we
model the absorption and quenching of light energy by photochemistry and
non-photochemical quenching, and the responses that occur (i.e.,
photoinhibition, photodamage, and symbiont loss) when these quenching
capacities are overwhelmed. While bleaching in response to high light
alone has been observed experimentally (Schutter et al. 2011; Downs et
al. 2013), mass coral bleaching events occur concurrently with high
temperature (Hoegh-Guldberg 1999); thus, it is important to justify our
consideration of light as the primary stressor. In reality, light and
temperature interact synergistically (Coles and Jokiel 1978; Jones et
al. 1998), and in fact, any stressor that disrupts the quenching of
light energy may lead to bleaching (Wooldridge 2010; Baker and Cunning
2015). This is because the proximate cause of photo-oxidative stress is
excess excitation energy, but the upstream events that lead to this
situation may be diverse: indeed, elevated temperature may inhibit
Rubisco functioning (Jones et al. 1998) and the repair of the D1 protein
in photosystem II (Warner, Fitt, and Schmidt 1999), which reduces the
capacity of photochemical quenching and leads to an excess of light
energy. In this way, elevated temperature serves to reduce the threshold
above which light stresses the system (Hoegh-Guldberg 1999);
importantly, light is still the proximate stressor. Therefore, we
omitted temperature from the model to maintain a desired level of
simplicity, while still allowing photooxidative stress and bleaching to
be simulated with biological realism in response to light.

Below we describe the formulation of each flux in the model.

@@

\begin{longtable}[]{@{}llll@{}}
\toprule
Symbol & Description & Units & Eq. no.\tabularnewline
\midrule
\endhead
\(j_X\) & Prey uptake rate & mol X C-mol H\textsuperscript{-1}
d\textsuperscript{-1} & 3\tabularnewline
\(j_N\) & Nitrogen uptake rate & mol N Cmol H\textsuperscript{-1}
d\textsuperscript{-1} & 4\tabularnewline
\(j_{HG}\) & Host biomass formation rate & C-mol H C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 5\tabularnewline
\(j_{HT}\) & Host biomass turnover rate & C-mol H C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 21\tabularnewline
\(r_{NH}\) & Recycled nitrogen from host turnover & mol N C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 6\tabularnewline
\(\rho_N\) & Nitrogen shared with the symbiont & mol N C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 7\tabularnewline
\(j_{eC}\) & Excess carbon used to activate host CCMs & mol C C-mol
H\textsuperscript{-1} d\textsuperscript{-1} & 8\tabularnewline
\(j_{CO_2}\) & CO\textsubscript{2} input to photosynthesis & mol
CO\textsubscript{2} C-mol H\textsuperscript{-1} d\textsuperscript{-1} &
9\tabularnewline
\(j_L\) & Light absorption rate & mol q C-mol S\textsuperscript{-1}
d\textsuperscript{-1} & 11\tabularnewline
\(r_{CH}\) & Recycled CO\textsubscript{2} from host turnover & mol
CO\textsubscript{2} C-mol H\textsuperscript{-1} d\textsuperscript{-1} &
12\tabularnewline
\(r_{CS}\) & Recycled CO\textsubscript{2} from symbiont turnover & mol
CO\textsubscript{2} C-mol S\textsuperscript{-1} d\textsuperscript{-1} &
13\tabularnewline
\(j_{CP}\) & Photosynthesis rate & mol C C-mol S\textsuperscript{-1}
d\textsuperscript{-1} & 14\tabularnewline
\(j_{eL}\) & Light energy in excess of photochemistry & mol q C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 15\tabularnewline
\(j_{NPQ}\) & Total capacity of NPQ & mol q C-mol S\textsuperscript{-1}
d\textsuperscript{-1} & 16\tabularnewline
\(c_{ROS}\) & ROS production proportional to baseline & -- &
17\tabularnewline
\(r_{NS}\) & Recycled nitrogen from symbiont turnover & mol N C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 18\tabularnewline
\(j_{SG}\) & Symbiont biomass formation rate & C-mol S C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 19\tabularnewline
\(\rho_C\) & Fixed carbon shared with host & mol C C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 20\tabularnewline
\(j_{ST}\) & Symbiont biomass turnover rate & C-mol S C-mol
S\textsuperscript{-1} d\textsuperscript{-1} & 21\tabularnewline
\bottomrule
\end{longtable}

Table 2. Model parameters (units explained in the text). Derivations of
each parameter value along with supporting references are provided in
the Supplementary Information.

@@

\end{minipage} & \begin{minipage}[b]{0.46\columnwidth}\raggedright\strut

@@

@@

\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut

@@

@@

mol C mol q\textsuperscript{-1}\strut

@@

mol ph C-mol S\textsuperscript{-1} d\textsuperscript{-1}\strut

@@

\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright\strut
Scaling parameter for strength of bleaching response\strut

@@

Additionally, the coral animal acquires nitrogen dissolved in the
surrounding seawater, which is assumed to represent ammonium, the
primary form utilized by corals (Yellowlees, Rees, and Leggat 2008).
This gives the host (rather than the symbiont) priority in nitrogen
utilization; this capacity is supported by experimental evidence (Wang
and Douglas 1998) and is consistent with the spatial arrangment of the
partners, where the host is in direct contact with the external
environment. The uptake of nitrogen from the environment is thus
specified by Michaelis-Menten kinetics using a maximum uptake rate
\(j_{Nm}\) and half-saturation constant \(K_N\):

@@

where \(\rho_C\) is fixed carbon shared by the symbiont (see Eq. 20),
and \(r_{NH}\) is recycled nitrogen liberated by host biomass turnover
(see Eq. XX). Host biomass turnover is equal to the specific maintenance
rate of host biomass, which is set as a parameter \(j_{HT}^0\),

@@

CO\textsubscript{2} arrives at the photosynthesis SU from multiple
sources: in addition to the CO\textsubscript{2} actively supplied by the
host through its CCMs (\(j_{CO_2}\); Eq. 9), we assume a fixed
proportion \(\sigma_{CH}\) of metabolic CO\textsubscript{2} generated by
host biomass turnover is passively available to the photosynthesis SU,
according to:

@@

where \(j_{CPm}\) is the maximum specific rate of photosynthesis, and
\(c_{ROS}\) is the relative rate of reactive oxygen species production
(see Eq. 17). Dividing the photosynthetic rate by \(c_{ROS}\) causes a
decline in response to photooxidative stress at high light levels, and
the emergent outcome of this SU formulation demonstrates a classic
photoinhibition response (Fig. S2).

@@

Carbon fixed by photosynthesis (\(j_{CP}\); Eq. 14) is then combined
with nitrogen shared by the host (\(\rho_N\); Eq. 7) and nitrogen
recycled from symbiont biomass turnover\footnote{Note that recycling of
  symbiont biomass turnover (\(rNS\) and \(rCS\)) only occurs based on
  the maintenance component of turnover (i.e., \(j_{ST}^0\)), and not
  the photodamage/bleaching component, as this loss represents biomass
  that is expelled from the host.}

@@

The model dynamics are specified by the differential equations (1-2)
that impose biomass balance for host and symbiont and by a set of
coupled non-linear algebraic equations (3-21) that define fluxes.
Several of these fluxes are defined \emph{implicitly}; for example, the
rejection fluxes of carbon and nitrogen from the symbiont and host
biomass SUs, respectively, act as reciprocal input fluxes to the other
SU. Similarly, the photosynthesis SU receives CO\textsubscript{2} at a
rate proportional to the carbon rejection flux from the host biomass SU,
and the rejection flux of excitation energy from the photosynthesis SU
acts to reduce its own production through photoinhibition. Without
further assumptions, however, the dynamical system is not always
unambiguously defined because for some combinations of parameters and
environmental forcing functions the system of algebraic equations has
more than one solution with all fluxes non-negative (see results below).
In such circumstances, the right hand side of the differential equations
(1) and (2) is not uniquely defined even when \(S\) and \(H\) are
specified. We resolved this problem by \emph{defining the dynamical
system} as the limit as a time step \(\Delta t \rightarrow 0\) of a
discretized system corresponding to Euler integration of the
differential equations, with those fluxes that represent flows of
elemental matter implemented by assuming that transfer of material
between components of the system takes one time step. Thus, for example,
CO\textsubscript{2} rejected from the host SU at time \(t\) arrives at
the photosynthesis SU at time \(t + \Delta t\).

@@

In a constant environment, the system ultimately reaches a steady state
of exponential growth or decline. However, under some conditions, either
of these outcomes may occur depending on initial values of symbiont and
host biomass, indicating the presence of alternate stable states (Fig.
2). The mechanism that produces these alternate stable states is the
positive feedback between carbon-limitation of the host and
CO\textsubscript{2}-limitation of photosynthesis: if symbiont biomass is
initially very low (i.e., a ``bleached'' coral), the system cannot
escape this positive feedback and cannot grow (unless feeding is
sufficiently high). However, if symbiont biomass is initially high
(i.e., a ``healthy'' coral), then the system remains in a
nitrogen-limited state with positive growth. For practical purposes,
this section of the manuscript considers only positive growth steady
states under constant environments; subsequently, we explore how
environmental forcing may cause the system to switch between alternate
stable states, which we interpret in the context of coral bleaching (see
``Coral Bleaching and Recovery'', below).

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig2.png}
\caption{Alternate stable states}
\end{figure}

@@

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-25 mol
photons m\textsuperscript{-2} s\textsuperscript{-1} light and
\textasciitilde{}4 µM DIN (Fig. 3A), are comparable to the rate of 0.07
d\textsuperscript{-1} measured by Tanaka et al. (2007) in \emph{Acropora
pulchra} under similar N-enriched conditions. Under conditions more
typical of reef environments (\textless{}0.5 µM DIN), predicted growth
rates are \textasciitilde{}0.01-0.03 d\textsuperscript{-1}. Observed
specific growth rates in several coral species fall near or below the
lower end of this range (\textasciitilde{}0.01 d\textsuperscript{-1})
(Osinga et al. 2011; Osinga et al. 2012), though values as high as 0.025
d\textsuperscript{-1} have been reported in \emph{Galaxea fascicularis}
(Schutter et al. 2010), and 0.04 d\textsuperscript{-1} in \emph{Aiptasia
diaphana}, a non-calcifying symbiotic anemone (Armoza-Zvuloni et al.
2014). It is not surprising that the model predicts slightly higher
growth than is typically observed, since it does not account for
ecological factors that may limit growth (e.g., competition, predation,
bioerosion). Furthermore, while most measurements are made on skeletal
growth, the model predicts biomass growth, which may not always be
strongly correlated (Anthony 2002).

>>>>

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  \emph{Develop a new module of photo-oxidative stress.} Of primary
  interest to coral biologists and ecologists is symbiosis dysfunction
  under environmental stress, resulting in coral ``bleaching''--the loss
  of algal symbionts from the association (Jokiel and Coles 1977).
  Bleaching is thought to be triggered by photo-oxidative stress in
  \emph{Symbiodinium} (Weis 2008), which has been modeled previously as
  a response to absolute external irradiance (Eynaud, Nisbet, and Muller
  2011). However, this response may also depend on self-shading by other
  symbionts (Enríquez, Méndez, and Iglesias-Prieto 2005),
  CO\textsubscript{2} availability at the site of photosynthesis
  (Wooldridge 2009), and other non-photochemical quenching mechanisms
  (Roth 2014). We incorporate these features into a new photo-oxidative
  stress module linking overreduction of the photosynthetic light
  reactions to downstream impacts of photoinhibition and photodamage.
  Importantly, this formulation introduces a link between
  CO\textsubscript{2}-limitation of photosynthesis and bleaching, and
  potential synergistic roles for heterotrophy and nutrient availability
  in influencing bleaching responses (Wooldridge 2014b)
\item
  \emph{Reduce theoretical and mathematical complexity.} Following the
  logic of Jager, Martin, and Zimmer (2013), we exclude certain features
  of formal DEB theory in order to capture behaviors of interest with
  the simplest possible formulation. Here, we present a model without
  reserves, maturity, or reproduction (see Kooijman 2010). This
  formulation precludes modeling the full life cycle of corals as
  reproduction, larval stages, and metamorphosis are not considered, but
  greatly reduces theoretical complexity and parameter numbers, which is
  advantageous given the relative paucity of data for corals. Moreover,
  dynamics of the symbiosis (i.e., changes in symbiont to host biomass,
  including bleaching and recovery) and coral biomass growth are
  efficiently captured with this simpler formulation, which also
  increases accessibility for biologists and ecologists without
  requiring significant expertise in DEB theory.
\item
  \emph{Provide well-documented, open-access code.} In order to
  facilitate the continued development and application of theoretical
  modeling tools for coral symbioses, we provide open access to the
  model in the form of detailed, commented code written in the R
  language (R Core Team 2014). With an accessible and modular framework,
  we envision this model as a resource for further development by the
  scientific community to include additional complexity and
  problem-specific components. We chose the R language, which is freely
  available and in common use by biologists and ecologists, to reach the
  widest possible audience with this work.
\end{enumerate}

With these as our primary motivations, we describe a simplified approach
to bioenergetic modeling of coral-\emph{Symbiodinium} symbioses that
dynamically integrates the influences of external irradiance, nutrients,
and prey availability on coral growth and symbiosis dynamics (i.e.,
changes in symbiont:host biomass ratios), allowing for the possibility
of coral bleaching in response to photooxidative stress. An emergent
finding of this work is that coral bleaching can be interpreted as an
alternate stable state of the symbiotic system, which provides a new
framework for understanding the mechanisms that drive a coral into a
bleached state, as well as those that facilitate recovery. In the
following sections, we describe and provide rationale for the model
structure, demonstrate a range of steady state and dynamic behaviors
that are consistent with observed phenomena, and discuss new insights
from this work in understanding responses of coral-\emph{Symbiodinium}
symbioses to environmental change.

@@

In this dynamical system, both the coral host and algal symbiont acquire
and use carbon and nitrogen to construct biomass. The symbiont fixes
carbon through photosynthesis and receives nitrogen shared by the host,
while the host acquires nitrogen from the environment and receives
carbon shared by the symbiont. A graphical representation of the model
is presented in Fig. 1, and each model flux and parameter is defined in
Tables 1 and 2, respectively. We use C-moles (C-mol) as the unit of
biomass for consistency with the rigorous mass balance of DEB theory: 1
C-mol is equivalent to the amount of biomass containing 1 mole of carbon
atoms. Host biomass (expressed as C-mol H), symbiont biomass (expressed
as C-mol S), and prey biomass (expressed as C-mol X) have fixed, but
different, molar N:C ratios (Table 2). Biomass is produced from carbon
and nitrogen by synthesizing units (SU), which are mathematical
specifications of the formation of a product from two substrates; we use
the ``parallel complementary'' formulation of Kooijman (2010) (page 105,
Fig. 3.7) to specify these fluxes. The two state variables of this
system are symbiont biomass and coral biomass; because resources are
acquired proportionally to surface area, and surface area is
proportional to volume, biomass increases exponentially during growth
(indeed, corals grow exponentially (Bak 1976)).

Environmental stress is implemented in the form of photooxidative
stress, which is thought to be a primary trigger of coral bleaching
(Lesser 1997; Weis 2008; Wooldridge 2009). To simulate bleaching, we
model the absorption and quenching of light energy by photochemistry and
non-photochemical quenching, and the responses that occur (i.e.,
photoinhibition, photodamage, and symbiont loss) when these quenching
capacities are overwhelmed. While bleaching in response to high light
alone has been observed experimentally (Schutter et al. 2011; Downs et
al. 2013), mass coral bleaching events occur concurrently with high
temperature (Hoegh-Guldberg 1999). Thus, it is important to justify our
consideration of light as the primary stressor. In reality, light and
temperature interact synergistically (Coles and Jokiel 1978; Jones et
al. 1998), and in fact, any stressor that disrupts the quenching of
light energy may lead to bleaching (Wooldridge 2010; Baker and Cunning
2015). This is because the proximate cause of photo-oxidative stress is
excess excitation energy, but the upstream events that lead to this
situation may be diverse: indeed, elevated temperature may inhibit
Rubisco functioning (Jones et al. 1998) and the repair of the D1 protein
in photosystem II (Warner, Fitt, and Schmidt 1999), which reduces the
capacity of photochemical quenching and leads to an excess of light
energy. In this way, elevated temperature serves to reduce the threshold
above which light stresses the system (Hoegh-Guldberg 1999);
importantly, light is still the proximate stressor. Therefore, we
omitted temperature from the model to maintain a desired level of
simplicity, while still allowing photooxidative stress and bleaching to
be simulated with biological realism in response to light.

@@

@@

\end{minipage} & \begin{minipage}[b]{0.44\columnwidth}\raggedright\strut
\end{minipage} & \begin{minipage}[b]{0.26\columnwidth}\raggedright\strut
Eq. no.\strut
\(j_X\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Prey uptake rate\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol X C-mol H\textsuperscript{-1} d\textsuperscript{-1}\strut
3\strut
\(j_N\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Nitrogen uptake rate\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol N Cmol H\textsuperscript{-1} d\textsuperscript{-1}\strut
4\strut
\(j_{HG}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Host biomass formation rate\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
5\strut
\(j_{HT}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Host biomass turnover rate\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
6\strut
\(r_{NH}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Recycled nitrogen from host turnover\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
7\strut
\(\rho_N\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Nitrogen shared with the symbiont\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
8\strut
\(j_{eC}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Excess carbon used to activate host CCMs\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol C C-mol H\textsuperscript{-1} d\textsuperscript{-1}\strut
9\strut
\(j_{CO_2}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
CO\textsubscript{2} input to photosynthesis\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol CO\textsubscript{2} C-mol H\textsuperscript{-1}
d\textsuperscript{-1}\strut
10\strut
\(j_L\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Light absorption rate\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol photons C-mol S\textsuperscript{-1} d\textsuperscript{-1}\strut
12\strut
\(r_{CH}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Recycled CO\textsubscript{2} from host turnover\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol CO\textsubscript{2} C-mol H\textsuperscript{-1}
d\textsuperscript{-1}\strut
13\strut
\(r_{CS}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Recycled CO\textsubscript{2} from symbiont turnover\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol CO\textsubscript{2} C-mol S\textsuperscript{-1}
d\textsuperscript{-1}\strut
14\strut
\(j_{CP}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Photosynthesis rate\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
15\strut
\(j_{eL}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Light energy in excess of photochemistry\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol photons C-mol S\textsuperscript{-1} d\textsuperscript{-1}\strut
16\strut
\(j_{NPQ}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Total capacity of NPQ\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol photons C-mol S\textsuperscript{-1} d\textsuperscript{-1}\strut
17\strut
\(c_{ROS}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
ROS production proportional to baseline\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
18\strut
\(r_{NS}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Recycled nitrogen from symbiont turnover\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
mol N C-mol S\textsuperscript{-1} d\textsuperscript{-1}\strut
19\strut
\(j_{SG}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Symbiont biomass formation rate\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
20\strut
\(\rho_C\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Fixed carbon shared with host\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
21\strut
\(j_{ST}\)\strut
\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Symbiont biomass turnover rate\strut
\end{minipage} & \begin{minipage}[t]{0.26\columnwidth}\raggedright\strut
22\strut

Table 2. Model parameters (units explained in the text). Justification
and/or derivation of each parameter value along with supporting
references are provided in the Supplementary Information.
\end{minipage} & \begin{minipage}[b]{0.44\columnwidth}\raggedright\strut

@@

\end{minipage} & \begin{minipage}[b]{0.12\columnwidth}\raggedright\strut

@@

\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut

@@

\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut

@@

mol C mol photons\textsuperscript{-1}\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut

@@

mol photons C-mol S\textsuperscript{-1} d\textsuperscript{-1}\strut
\end{minipage} & \begin{minipage}[t]{0.12\columnwidth}\raggedright\strut

@@

\end{minipage} & \begin{minipage}[t]{0.44\columnwidth}\raggedright\strut
Scaling parameter for bleaching response\strut

@@

Additionally, the coral animal acquires dissolved inorganic nitrogen
(DIN) from the surrounding seawater, which is assumed to represent
ammonium, the primary form utilized by corals (Yellowlees, Rees, and
Leggat 2008). This gives the host (rather than the symbiont) priority in
nitrogen utilization; this capacity is supported by experimental
evidence (Wang and Douglas 1998) and is consistent with the spatial
arrangment of the partners, where the host is in direct contact with the
external environment. The uptake of nitrogen from the environment is
thus specified by Michaelis-Menten kinetics using a maximum uptake rate
\(j_{Nm}\) and half-saturation constant \(K_N\):

@@

where \(\rho_C\) is fixed carbon shared by the symbiont (see Eq. 21),
and \(r_{NH}\) is recycled nitrogen liberated by host biomass turnover
(see Eq. 7). Host biomass turnover is equal to the specific maintenance
rate of host biomass,

@@

CO\textsubscript{2} arrives at the photosynthesis SU from multiple
sources: in addition to the CO\textsubscript{2} actively supplied by the
host through its CCMs (\(j_{CO_2}\); Eq. 10), we assume a fixed
proportion \(\sigma_{CH}\) of metabolic CO\textsubscript{2} generated by
host biomass turnover is passively available to the photosynthesis SU,
according to:

@@

where \(j_{CPm}\) is the maximum specific rate of photosynthesis, and
\(c_{ROS}\) is the relative rate of reactive oxygen species production
(see Eq. 18). Dividing the photosynthetic rate by \(c_{ROS}\) causes a
decline in response to photooxidative stress at high light levels, and
the emergent outcome of this SU formulation demonstrates a classic
photoinhibition response (Fig. S2).

@@

Carbon fixed by photosynthesis (\(j_{CP}\); Eq. 15) is then combined
with nitrogen shared by the host (\(\rho_N\); Eq. 8) and nitrogen
recycled from symbiont biomass turnover\footnote{See footnote 2.}

@@

The model dynamics are specified by the differential equations (1-2)
that impose biomass balance for host and symbiont and by a set of
coupled non-linear algebraic equations (3-22) that define fluxes.
Several of these fluxes are defined \emph{implicitly}; for example, the
rejection fluxes of carbon and nitrogen from the symbiont and host
biomass SUs, respectively, act as reciprocal input fluxes to the other
SU. Similarly, the photosynthesis SU receives CO\textsubscript{2} at a
rate proportional to the carbon rejection flux from the host biomass SU,
and the rejection flux of excitation energy from the photosynthesis SU
acts to reduce its own production through photoinhibition. Without
further assumptions, however, the dynamical system is not always
unambiguously defined because for some combinations of parameters and
environmental forcing functions the system of algebraic equations has
more than one solution with all fluxes non-negative (see results below).
In such circumstances, the right hand side of the differential equations
(1) and (2) is not uniquely defined even when \(S\) and \(H\) are
specified. We resolved this problem by \emph{defining the dynamical
system} as the limit as a time step \(\Delta t \rightarrow 0\) of a
discretized system corresponding to Euler integration of the
differential equations, with those fluxes that represent flows of
elemental matter implemented by assuming that transfer of material
between components of the system takes one time step. Thus, for example,
CO\textsubscript{2} rejected from the host SU at time \(t\) arrives at
the photosynthesis SU at time \(t + \Delta t\).

@@

In a constant environment, the system ultimately reaches a steady state
of exponential growth or decline. However, under some conditions, either
of these outcomes may occur depending on initial values of symbiont and
host biomass, indicating the presence of alternate stable states (Fig.
2). The mechanism that produces these alternate stable states is the
positive feedback between carbon-limitation of the host and
CO\textsubscript{2}-limitation of photosynthesis: if symbiont biomass is
initially very low (i.e., a ``bleached'' coral), very little carbon is
fixed, and the system cannot escape this positive feedback and cannot
grow (unless feeding is sufficiently high). However, if symbiont biomass
is initially high (i.e., a ``healthy'' coral), then the system remains
in a nitrogen-limited state with positive growth. For practical
purposes, this section of the manuscript considers only positive growth
steady states under constant environments; subsequently, we explore how
environmental forcing may cause the system to switch between alternate
stable states, which we interpret in the context of coral bleaching (see
``Coral Bleaching and Recovery'', below).

\begin{figure}[htbp]
\centering
\includegraphics{../img/Fig2.png}
\caption{Alternate stable states in S:H biomass and growth across a
light gradient. Alternate stable states occur between
\textasciitilde{}33-44 mol photons m\textsuperscript{-2}
d\textsuperscript{-1} under these conditions (DIN=1e-7 molN
L\textsuperscript{-1}; X=2e-6 molX L\textsuperscript{-1}), depending on
whether initial S:H is high (0.1, closed circles), representing a
healthy coral, or low (0.0001, open circles), representing a bleached
coral.}
\end{figure}

@@

The maximum predicted growth rates of \textasciitilde{}0.1
d\textsuperscript{-1}, occurring between \textasciitilde{}10-25 mol
photons m\textsuperscript{-2} s\textsuperscript{-1} light and
\textasciitilde{}4 µM DIN (Fig. 3A), are comparable to the rate of 0.07
d\textsuperscript{-1} measured by Tanaka et al. (2007) in \emph{Acropora
pulchra} under similar N-enriched conditions. Under conditions more
typical of reef environments (\textless{}0.5 µM DIN), predicted growth
rates are \textasciitilde{}0.01-0.03 d\textsuperscript{-1}. Observed
specific growth rates in several coral species fall near or below the
lower end of this range (\textasciitilde{}0.01 d\textsuperscript{-1})
(Osinga et al. 2011; Osinga et al. 2012), though values as high as 0.025
d\textsuperscript{-1} have been reported in \emph{Galaxea fascicularis}
(Schutter et al. 2010), and 0.04 d\textsuperscript{-1} in \emph{Aiptasia
diaphana}, a non-calcifying symbiotic anemone (Armoza-Zvuloni et al.
2014). However, it is not surprising that observed growth rates are
often lower than model predictions, since the model does not account for
ecological factors that may limit growth (e.g., competition, predation,
bioerosion). Furthermore, while most measurements are made on skeletal
growth, the model predicts biomass growth, which may not always be
strongly correlated (Anthony 2002).

At irradiance levels above \textasciitilde{}25 mol photons
m\textsuperscript{-2} d\textsuperscript{-1}, steady state growth rates
decline until positive growth ceases above \textasciitilde{}40 µmol
photons m\textsuperscript{-2} d\textsuperscript{-1} (Fig. 3A). The
mechanism underlying this decline is the increase in light energy beyond
the capacities of photosynthesis and non-photochemical quenching: excess
excitation energy generates reactive oxygen species (ROS) (Weis 2008;
Roth 2014), which, in this model, have the phenomenological consequences
of reducing the photosynthetic rate (representing photoinhibition) and
increasing symbiont biomass loss (representing photodamage and/or
symbiont expulsion) (see Eynaud, Nisbet, and Muller 2011). Together,
these impacts reduce the symbiont to host biomass ratio (Fig. 3B), as
occurs during coral bleaching. This reduction in symbionts consequently
reduces the flux of fixed carbon to the host, resulting in increasing
carbon-limitation (Fig. 3B) and eventual cessation of growth (Fig. 3A).
